---
title: "MM23"
output: html_document
date: "2023-12-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(ggplot2)
library(dplyr)
library(lme4)
library(LabApplStat)
library(readr)
library(gridExtra)
library(pbkrtest)

load("/home/asr/Desktop/statA/ass1/DWBdata.Rdata")
fit1 <- lmer(DWB100 ~ WkDay + z2sleep + (z2sleep|NameID), data=DWBdata)
```

# 1.

We test $H_0 EY \in L_{z2sleep}$, the hypothesis is that weekday has no effect on daily well being. We first do an F-test based on SatterWhathes method.
```{r}
null_fit <- lmer(DWB100 ~ z2sleep + (z2sleep|NameID), data=DWBdata)
SATmodcomp(fit1, null_fit)
```
And get a very small p-value. We do simulation and chi-sq approximated testing
```{r}
PBmodcomp(fit1, null_fit)
```
And also get a very small p-value in both cases. The p-values are comparable.

# 2.

Consider the hypothesis that daily well-being is the same on sundays and saturdays. We can modify the data, to have the same level for sunday and saturday. An F-test gives
```{r}
DWBdata <- DWBdata %>% mutate(wk_day_mod = factor(case_when(as.numeric(WkDay) >= 6 ~ as.factor(6),
                                          TRUE ~ WkDay)))
null_fit <- lmer(DWB100 ~ wk_day_mod + +z2sleep + (z2sleep|NameID), data=DWBdata)
SATmodcomp(fit1, null_fit)
```
Which is not significant. We do simulation and LRT
```{r}
PBmodcomp(fit1, null_fit)
```
Which also gives large p-values. An alternative is to set the contrast to sunday and do a t-test
```{r}
DWBdata$WkDay <- factor(DWBdata$WkDay, levels = c(7,seq(1,6)))
fit1 <- lmerTest::lmer(DWB100 ~ WkDay + z2sleep + (z2sleep|NameID), data=DWBdata)
fit1 %>% summary()
```
Which also gives the same result.

## 3.

We can test for differenes between weekend/weekdays the same as above

```{r}
DWBdata <- DWBdata %>% mutate(weekend = factor(case_when(as.numeric(WkDay) >= 6 ~ "weekend",
                                          TRUE ~ "workday")))
null_fit <- lmer(DWB100 ~ weekend + +z2sleep + (z2sleep|NameID), data=DWBdata)
SATmodcomp(fit1, null_fit)

```
Which is very much significant.
